A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-06-28 rigid_bodies.tex
Tentativa de corrigir git pull
Mudanças na introdução

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
\mathchardef\mhyphen="2D
\usepackage{booktabs}
\usepackage{multirow}
\begin{document}

@@

One of the strategies commonly used to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Widely used iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieve

the dynamics of rigid bodies is able to keep this type of constraint implicitly, and thus emerges as a suitable approach in these cases.

one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient. In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble. The property of being symplectic implies the preservation of phase space volume. This feature, together with time reversibility are mandatory for numerical integrators used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}, etc. It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.

In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform MD. The contribution to the NVE case relies on a factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}. In the NVT case, both the translational and rotational degrees of freedom (DOF) are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF. We have also simplified the calculation of the pressure for a rigid body system, whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions (PBC). The expression we derive depends on the resultant interaction forces between rigid body pairs. This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces.

This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy. Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy. Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
\mathchardef\mhyphen="2D
\usepackage{booktabs}
\usepackage{multirow}

\begin{document}

@@

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach. It is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are mandatory properties for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators. The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two independent thermostat chains act separately on the rotational and the translational degrees of freedom. We have also simplified the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions (PBC). The expression we derive depends on the resultant interaction forces between body pairs. This avoids the evaluation of constraint forces, which are necessary if the pressure is calculated in terms of the interatomic forces.

This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy. Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy. Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.
